\bf Model Visualization Techniques for a Social Network Model

### 2017-06-02 paper.Rnw
up to bottom of page 10

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \citep{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) 
@
 \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$. We denote these network observations $x(t_1), \dots, x(t_M)$, and in modelling we condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), call it $X(T)$. This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$. All other parts of the process are unseen. The process $X(T)$ is a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie. Once an actor is given the chance to change a tie, it tries to maximize a sort of utility function based on the current and potential future states of the network.

@@

For the network $x$ and each actor $i$ in the network, the rate function dictates how often the actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network. For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. In other modelling scenarios, this rate can be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$. We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$. This assumption leads to the rate for the whole network, $n\alpha$. In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$

@@

Thanks to the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is written as $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p. 371]{snijders01}. There are many possible parameters $\beta_i$ to add to the model. They can be split up into two groups: first, the structural effects, which only depend on the structure of the network. The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$. The second set of effects are the actor-level or covariate effects. These effects also depend on the structure of the network. They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can be included in the objective function, see \citet{RSiena}.

>>>>

There are two primary methods used to visualize networks: node-link diagrams and matrix visualization \cite{knuth2000, adjmatpro}. As a toy example, let us assume that we have five nodes, $\{1,2,3,4,5\}$, connected by five edges:  $\{2 \mapto 4, 3 \mapto 4, 1 \mapto 5, 3 \mapto 5, 5 \mapto 4\}$ We use this toy data set to demonstrate the two visualization methods in Figure~\ref{fig:toyplots}.

@@

ggplot(data = toynet$edges) +
  geom_tile(aes(x = from, y = to), color = 'grey20', fill = 'black') +
  geom_tile(aes(x = to, y = from), color = 'grey20', fill = 'black') +
  geom_segment(aes(x = x, xend = xend), y = 0.5, yend=5.5, colour="grey30", data = data.frame(x = 0:5+.5, xend= 0:5+.5)) + 
  geom_segment(aes(y=y, yend=yend), x = 0.5, xend=5.5, colour="grey30", data = data.frame(y = 0:5+.5, yend= 0:5+.5)) + 
  scale_x_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  scale_y_continuous(breaks = 1:5, labels = 1:5, limits = c(0.5,5.5)) +
  theme_bw() + 
  theme(panel.grid = element_blank()) +
  labs(title = "Adjacency Matrix Visualization") + 
  theme(aspect.ratio = 1) +
  xlab("from (ego)") +
  ylab("to (alter)")
@
 \hh{XXX Explain each of these visualizations and either comment on the differences in the visualizations (colors) or don't show them. }

@@

In this paper, the term \textit{dynamic network} refers to a network consisting of the same set of $n$ nodes that is changing over time, and is observed at $M$ discrete time points, $t_1, \dots, t_M$ with $t_1 \le t_2 \le \dots \le t_M$. We denote the network observation at timepoint $t_k$ by $x(t_k)$. In the modelling process we usually condition on the first observation, $x(t_1)$. The SAOM assumes that this longitudinal network is embedded within a continuous time Markov chain (CTMC), which we will refer to as $X(T)$. This process is almost entirely unobserved: assume that the beginning of the process, $X(0)$ is equivalent to the first network observation $x(t_1)$, while the end of the process $X(\infty)$ is equivalent to the last observation $x(t_M)$. All other parts of the process are unseen. We model the process $X(T)$ as a series of single tie (or edge) changes, in which one actor at a time is given the opportunity to add or remove one outgoing tie. Once an actor is given the chance to change a tie, it tries to maximize a utility function based on the current and potential future states of the network.

@@

For the network $x$ and each actor $i$ in the network, the number of times that an actor $i$ gets to change its ties, $x_{ij}$, to other nodes $j \neq i$ in the network is dictated by a \emph{rate function} $\alpha$. For this paper, we assume a simple rate function, $\alpha_m$ that is constant for all actors between observations at time $t_m$ to $t_{m+1}$. \hh{XXX that means that the function is essentially just a rate parameter, right?}
Other modelling scenarios allow this rate to be more flexible,  e.g.\  be a function that depends on the time period of observation, some actor-level covariates or some actor-level network statistics. 
\hh{How is the rate parameter different from a rate parameter?}
The rate parameter determines how quickly actor $i$ gets an opportunity to change one of its ties, $x_{ij}$ in the time period $t_{m} \leq T < t_{m+1}$. We assume that the actors $i$ are conditionally independent given their current ties, $x_{i1}, \dots, x_{in}$. This assumption leads to the rate for the whole network, $n\alpha$. In order to achieve the memorylessness property of a Markov process, for any time point, $T$, where $t_{m} \leq T < t_{m+1}$, the waiting time to the next change opportunity by actor $i$ is exponentially distributed with expected value $\alpha_m^{-1}$. Thus, the waiting time to the next change opportunity by \textit{any} actor in the network is also exponentially distributed with expected value $(n\alpha_m)^{-1}$

@@

\hh{shouldn't that be independence? - make the conditional independence assumption above a labelled formula (e.g. C1) and then call it here }
Because of the conditional dependence assumptions in the model, we can consider the objective function for each node separately, since only one tie from one node is changing at a time. The objective function is then written as \hh{the sum of the individual changes XXX write out the math in words XXX} $$f_i(\boldsymbol{\beta}, x) = \sum_k \beta_k s_{ik}(x, \mathbf{Z}),$$
for $x \in \mathcal{X}$, the space of all possible directed networks with the $n$ nodes, and $\mathbf{Z}$ matrix of covariates. The vector $\boldsymbol{\beta}$ are the parameters of the model with corresponding network and covariate statistics, $s_{ik}(x, \mathbf{Z})$, for $k = 1,\dots, K$. \hh{XXX ego node isn't defined yet - the description of the charts would be a good place for it.} Given the ego node, $i$, there are $n$ possible steps for the actor $i$ to take: either one of all current ties $x_{ij} = 1$ will be destroyed, a new tie will be created, or no change will occur.

The parameters, $\boldsymbol{\beta}$, are attached to various actor-level network statistics, $s_{ik}(x)$. \hh{'There are always' ... that seems to come from a Snijder paper? Re-phrase this to be less vague: In the current implementation of RSiena (snijder) a model has minimally two parameters...} There are always at least two parameters, $\beta_1$ for the outdegree of a node, and $\beta_2$ for the number of reciprocal ties held by a node \citep[p. 371]{snijders01}. \hh{'There are many' another vague phrase} There are many possible parameters $\beta_i$ to add to the model. \hh{'They' ... be specific.} They can be split up into two groups: first, the structural effects, which only depend on the structure of the network. \sct{The inclusion of these effects has origin in the classical exponential random graph model (ERGM) for networks. These effects are written in terms of the edge variables $x_{ij}$, for $i \neq j$. The second set of effects are the actor-level or covariate effects.} \hh{XXX move the last colored section to the front of this paragraph. That sets up the structure nicely, e.g. `In determining the covariates of a SAOM, two sources of parameters are considered: ...} These effects also depend on the structure of the network. They are written in terms of $x_{ij}$ but also in terms of the covariates, $\mathbf{Z}$. A table of some possible structural and covariate effects is given in Table~\ref{tab:effects}. For a complete list of the network and covariate statistics that can be included \hh{currently} in the objective function, see \citet{RSiena}.
